Customizing Software by Direct Manipulation of Tabular Data

### 2020-05-22 paper.tex
first final

In this work, we ask: what would it look like to build a direct
maniulation interface that lowers the threshold for starting to
customize software? We take inspiration from spreadsheets and visual
database query interfaces \citep{bakke2016, 2020a}, which have
successfully enabled millions of end users to compute with data through
direct manipulation.

@@

Despite these limitations, we found that Wildcard does offer
predictability: once we built a site adapter for a website, it was
generally obvious what customizations were possible, and the UI guided
us towards building customizations that matched the system's model. If
data is not available in the table, for example, then it's abundantly
clear that customizations relying on that data will not be expressible.

We consider all of these limitations to be an acceptable outcome. Our
goal is to support as many useful customizations as possible with a low
threshold of difficulty, and not to span all possible customizations. If
users want to implement more sophisticated customizations, they have the
option of graduating to more advanced customization tools.

@@

@@

Some customization environments like Mac Automator and Zapier forego
textual syntax and let the user connect programs and construct
automations by dragging and dropping icons representing commands. These
environments still do not constitute direct manipulation, though: the
objects being manipulated are in the domain of programming, not in the
domain of the task at hand. Imperative programming is a reasonable
choice as the model for building customizations. Turing-complete
programming provides a high ceiling for possible customizations, and a
sequence of commands is a natural fit for automations that simulate a
series of steps taken by the user.

Imperative programming is a reasonable choice as the model for building
customizations. Turing-complete programming provides a high ceiling for
possible customizations, and a sequence of commands is a natural fit for
automations which simulate a series of steps taken by the user.

There is, however, a serious drawback to this approach. MacLean et al
\citep{maclean1990} describe an ideal for user-tailorable systems: a
``gentle slope'' from using to customizing, where small incremental
increases in skill lead to corresponding increments of customization
power. Requiring users wanting to customize their applications to learn
programming creates an abrupt ``cliff,'' exacting a significant
investment in learning even to implement the simplest customizations.
Another goal of MacLean et al is to make it ``as easy to change the
environment as it is to use it''---at least for some subset of changes.
But in scripting languages, even user-friendly ones like Applescript or
Chickenfoot, the experience of customization does not remotely resemble
the experience of use, so these systems can't meet this goal.

With table-driven customization we aim to provide a gentler slope, by
using direct manipulation for software customization. The data shown in
the table view is the domain data from the original application. The
user makes changes to the data by selecting areas of interest in the
table, e.g.~sorting/filtering by clicking the relevant column header, or
adding annotation by clicking on the relevant row. These interactions
are common in GUI applications, and Wildcard therefore meets MacLean et
al's goal: some one-click customizations are as easy as using the
original application.

One aspect of directness that we have chosen not to maintain in Wildcard
is enabling customization in the context of the original user interface,
as explored by tools like Scotty \citep{eagan2011}. We have found that
augmenting the original UI with a distinct, additional representation
provides a consistent experience across all applications, and clearly
shows the user what structured data is available to work with.

Ainsworth et al provide a helpful taxonomy of the value of multiple
representations \citep{ainsworth1999}. In their terms, Wildcard plays a
complementary role by supporting a different set of tasks from the
original application, while displaying shared information. Wildcard may
also help construct deeper understanding by subtraction. By stripping
away details and only showing the essential data in an interface,
Wildcard helps users think of new ways of using that data, outside the
specific constraints of the original application.

It is important to note that there are many specific customizations that
can be achieved in systems like Chickenfoot \citep{bolin2005} that
cannot be reproduced in Wildcard. We consider this an acceptable
tradeoff in exchange for a gentler slope in customization, and we show
in Section~\ref{sec:example} and Section~\ref{sec:reflections} that our
model can still implement many useful customizations in practice. Also,
there is sometimes a way to reframe an imperative script in terms of our
direct manipulation model; for example, a script that iterates through
rows in a page adding some additional information can be reproduced
using a formula in Wildcard.

@@

Software customization tools typically fall into one of two categories:
ad hoc, or semantic.

\emph{Ad hoc customization tools} enable customization without using
official extension APIs, enabling a broader range of customizations on
top of more applications. For example, web browser extensions have
demonstrated the utility of customizing websites through manipulating
the DOM, without needing explicit extension APIs to be built in.
However, ad hoc customization comes with a corresponding cost: these
tools can typically only operate at a low level of abstraction,
e.g.~manipulating user interface elements. This makes it harder for end
users to write scripts, and makes the resulting scripts more brittle.

\emph{Anticipated customization tools}, in contrast, use explicit
extension APIs provided by the application developer. Examples of this
include accessing a backend web API, or writing a customization in
Applescript that uses an application-specific API. The main benefit is
that this allows the extension author to work with meaningful concepts
in the application domain---``create a new calendar event'' rather than
``click the button that says new event.''---which makes customizations
easier to build and more robust. However, this style limits the range of
extensions that can be built to those that the official plugin API
supports.

\emph{todo: copy over more content from gdoc}

With Wildcard, we use a hybrid approach that aims to provide the best of
both worlds. Programmers implement an API wrapper that is internally
implemented as an ad hoc customization, but externally provides a
high-level interface to the application, abstracting away the details of
the user interface. These wrappers are added to a shared repository,
available to all users of the system. When an end user is using a site
that already has an adapter, they get a semantic customization
experience that avoids low-level details.

@@

With Wildcard we propose a decoupling of these two layers: a repository
of shared wrappers maintained by programmers, and a separate repository
of specific customizations built on top of these wrappers. This general
architecture has been successfully demonstrated by projects like
Gmail.js, an open source project that creates a convenient API for
browser extensions to interface with the Gmail web email client.

(\emph{todo: mention automated wrapper induction here or somewhere
else?})

@@

This paper extends work reported in a workshop paper by Litt and Jackson
\citep{litt2020} which presented an early prototype version of Wildcard.
We have substantially extended their work in this paper by creating the
table adapter abstraction and reimplementing the system around that
abstraction, evaluating the system more fully on many more websites, and
by characterizing the design of the system in more detail.

>>>>

In this work, we ask: what would it look like to build a direct
maniulation interface that lowers the threshold for starting to
customize software? We take inspiration from spreadsheets and visual
database query interfaces, which have successfully enabled millions of
end users to compute with data through direct manipulation.

@@

We consider these limitations to be an acceptable outcome. Our goal is
to support as many useful customizations as possible with a low
threshold of difficulty, and not to span all possible customizations. If
users want to implement more sophisticated customizations, they have the
option of graduating to more advanced customization tools like scripting
languages.

We have found that one benefit of showing structured data is
predictability: once we build an adapter for a website, it is clear what
data is available or unavailable for use in customizations. Also, there
is sometimes a way to reframe an imperative script in terms of our
direct manipulation model. For example, a script that iterates through
rows in a page adding some additional information to each row can be
reproduced using a single formula in Wildcard.

@@

Here we discuss some themes that we've explored with this work, which go
beyond our specific approach and address issues in software
customization more broadly.

@@

Some customization environments like Mac Automator and Zapier forego
textual syntax and let the user connect programs and construct
automations by dragging and dropping icons representing commands. These
environments still do not constitute direct manipulation, though: the
objects being manipulated are in the domain of programming, not in the
domain of the task at hand.

Imperative programming is a reasonable choice as the model for building
customizations. Turing-complete programming provides a high ceiling for
possible customizations, and a sequence of commands is a natural fit for
automations that simulate a series of steps taken by the user. There is,
however, a serious drawback to this approach. MacLean et al
\citep{maclean1990} describe an ideal for user-tailorable systems: a
``gentle slope'' from using to customizing, where small incremental
increases in skill lead to corresponding increments of customization
power. Requiring users wanting to customize their applications to learn
programming creates an abrupt ``cliff,'' exacting a significant
investment in learning even to implement the simplest customizations.
Another goal of MacLean et al is to make it ``as easy to change the
environment as it is to use it''---at least for some subset of changes.
But in scripting languages, the experience of customization does not
remotely resemble the experience of use.

With table-driven customization we aim to provide a gentler slope, by
using direct manipulation for software customization. The data shown in
the table view is the domain data from the original application. The
user makes changes to the data by selecting areas of interest in the
table, e.g.~sorting/filtering by clicking the relevant column header, or
adding annotations by clicking on the relevant row. These interactions
are common in GUI applications, and Wildcard therefore meets MacLean et
al's goal: some one-click customizations are as easy as using the
original application. Formulas introduce some additional complexity, but
the pure functional semantics and the intermediate visibility of results
make the task still easier than imperative programming.

One aspect of directness that we have chosen not to maintain in Wildcard
is enabling customization in closer proximity to the original user
interface elements, as explored by tools like Scotty \citep{eagan2011}.
We have found that augmenting the original UI with a distinct,
additional representation provides a more consistent experience across
all applications, and clearly shows the user what structured data is
available to work with.

Ainsworth et al provide a helpful taxonomy of the value of multiple
representations \citep{ainsworth1999}. In their terms, Wildcard plays a
\emph{complementary role} by supporting a \emph{different set of tasks}
from the original application, while displaying \emph{shared
information}. Wildcard may also help construct \emph{deeper
understanding by subtraction}. By stripping away details and only
showing the essential data in an interface, Wildcard helps users think
of new ways of using that data, outside the specific constraints of the
original application.

As noted in Section~\ref{sec:reflections}, there are many customizations
that can be achieved in scripting languages that cannot be implemented
in Wildcard. We consider this an acceptable tradeoff in exchange for a
gentler slope in customization, and we show in Section~\ref{sec:example}
and Section~\ref{sec:reflections} that our model can still implement
many useful customizations.

@@

\emph{Ad hoc customization tools} enable customization without using
official extension APIs, enabling a broader range of customizations on
top of more applications. For example, web browser extensions have
demonstrated the utility of customizing websites through manipulating
the DOM, without websites needing to provide explicit extension APIs.
However, ad hoc customization comes with a cost: these tools typically
operate at a low level of abstraction, e.g.~manipulating user interface
elements, rather than in a meaningful domain model. This makes it harder
for end users to write scripts, and makes the resulting scripts more
brittle as the specifics of a user interface change.

\emph{Anticipated customization tools}, in contrast, use explicit
extension APIs provided by the application developer. Examples of this
include accessing a backend web API, or writing a customization in
Applescript for an application that exposes its domain model to the
scripting language. The main benefit of this style is that it allows the
extension author to work with meaningful concepts in the application
domain---``create a new calendar event'' rather than ``click the button
that contains the text `new event'.''---which makes customizations
easier to build and more robust. However, the plugin API limits the
types of customizations that can be built, and many applications don't
have an extension system at all.

With Wildcard, we use a hybrid approach that aims to provide the best of
both worlds. Programmers implement site-specific adapters that are
internally implemented ad hoc customizations, but externally provide a
high-level interface to the application, abstracting away the details of
the user interface. These wrappers are added to a shared repository,
available to all users of the system. When an end user is using a site
that already has an adapter implemented, they benefit from a semantic
customization experience that avoids low-level details.

@@

With Wildcard we propose a decoupling of these two layers: a repository
of shared wrappers maintained by programmers, and a separate repository
of specific customizations built on top of these wrappers. This general
architecture has been successfully demonstrated by projects like
\href{https://github.com/KartikTalwar/gmail.js/}{Gmail.js}, an open
source project that creates a convenient API for browser extensions to
interface with the Gmail web email client.

The success of semantic wrappers depends on a key hypothesis: that a
single wrapper created by a programmer can be used for many different
purposes by end users. Although we've validated that a single generic
adapter can support many customizations, so far the people making the
adapters have largely been the same people building customizations on
top of them. One helpful future addition would be to allow end users to
create wrappers themselves, or at least to modify existing ones. There
are many existing projects exploring end-user web scraping (such as
Helena \citep{chasins2018}) which might prove helpful for this.

The distribution mechanism for semantic wrappers is also important.
Currently, the distribution mechanism is simply merging the code for all
adapters into the main Wildcard codebase. This is the simplest workable
solution, but has its downsides: contributing involves a fairly high
barrier of creating a pull request on Github, and using newly
contributed wrappers requires installing a new version of the extension.
In the future we might explore other mechanisms, like an online
repository that the extension pulls from dynamically. Security is also a
consideration---DOM scraping adapters can execute arbitrary Javascript
code, which opens up the possibility of malicious adapters being
contributed. Currently we plan to solve this challenge with centralized
code review, but another approach we are considering is using or
inventing a more restricted domain-specific language for specifying
scraping logic.

@@

This paper extends work reported in a workshop paper by Litt and Jackson
\citep{litt2020} which presented an earlier version of the Wildcard
extension. We have substantially extended their work in this paper by
creating the table adapter abstarction, reimplementing the internals of
Wildcard around that abstraction, evaluating the system on many more
websites, and by characterizing the design of the system in much more
detail than in their workshop paper.
